---
title: "MitoMicroPub_FinalDataAnalysis"
author: "Abigail Wendland"
date: "2026-04-25"
output: html_document
---

## Overview

- This analysis investigates mitochondrial health using red:green fluorescence ratios across different age groups and treatment conditions.
- the data used for analysis was obtaned using Image J and z-stacks gotten from the Zesis confocal. 


## Analysis Plan

1. Import and clean raw data sets 
2. Perform Statistical Tests (t-tests and )
3. Generate plots comapring age and treatment groups 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Install packages 

```{r packages used for analysis}
 install.packages(c("tidyverse" , "ggpubr" , "janitor" , "readxl" , "knitr" , "rmarkdown","here", "patchwork", "ggbeeswarm"))
```
### Load packages 

```{r load packages}
library(tidyverse)
library(ggpubr)
library(rstatix)
library(janitor)
library(readxl)
library(knitr)
library(here)
library(patchwork)
library(ggbeeswarm)
```
```{r}
dir.create("figures", showWarnings = FALSE)
dir.create("results", showWarnings = FALSE)
```

## Data Analysis for Ageing nSyb and Repo 

### Upload Data
``
```{r uplad data set}
data<- read_excel("~/Documents/MicroPub/s1_mito_timer_data_blinded.xlsx")
head(data)
```

### Clean Names
```{r clean data set}

data <- data %>% 
  clean_names() 
```
```{r}
data <- data %>% 
  mutate(treatment = case_when(
      grepl("PositiveControl", summary) ~ "paraquat",
      grepl("NegativeControl", summary) ~ "control",
      TRUE ~ "baseline"
    ),
    
    age = case_when(
      grepl("Young", summary) ~ "young",
      grepl("Middle", summary) ~ "middle",
      grepl("Old", summary) ~ "old",
      grepl("10to15", summary) ~ "10to15",
      grepl("15to18", summary) ~ "15to18"
    )
  )
```
```{r}
data <- data %>%
  mutate(
    treatment = factor(treatment),
    age = factor(age)
  )
```
```{r}

table(data$treatment, data$age)

```
```{r check data set}

data_long <- data %>%
  pivot_longer(
    cols = c(left_red_green_ratio, right_red_green_ratio),  
    names_to = "roi_side",
    values_to = "value"
  )
data_long <- data_long %>%
  mutate(
    value = as.numeric(value)
  )
```

### Plotting Data
```{r create datasets for nSyb and Repo red:green across age groups}
nsyb_data <- plot_data %>%
  filter(cell_type == "nSybGa4") %>%
  mutate(age = factor(age_only, levels = c("Young", "Middle", "Old")))

repo_data <- plot_data %>%
  filter(cell_type == "RepoGa4") %>%
  mutate(age = factor(age_only, levels = c("Young", "Middle", "Old")))

```
```{r run stats on nSyb and Repo red:green across age groups}
library(rstatix)

run_age_stats <- function(df, group_name) {
  
  anova_out <- df %>%
    anova_test(value ~ age)
  
  tukey_out <- df %>%
    tukey_hsd(value ~ age) %>%
    add_xy_position(x = "age")
  
  anova_save <- anova_out %>%
    select(where(~ !is.list(.)))
  
  tukey_save <- tukey_out %>%
    select(where(~ !is.list(.)))
  
  write.csv(
    anova_save,
    paste0("results/", group_name, "_age_anova.csv"),
    row.names = FALSE
  )
  
  write.csv(
    tukey_save,
    paste0("results/", group_name, "_age_tukey.csv"),
    row.names = FALSE
  )
  
  return(tukey_out)
}


```


```{r create plots for nSyb red:green aging data}

stat_pvalue_manual(
  repo_stats,
  label = "p.adj.signif",
  tip.length = 0.01
)

p_nsyb <- ggplot(nsyb_data, aes(x = age, y = value, color = age)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, color = "gray50") +
  geom_jitter(width = 0.05, height = 0, size = 2.5, alpha = 0.65) +
  stat_pvalue_manual(
    nsyb_stats,
    label = "p.adj.signif",
    tip.length = 0.01
  ) +
  scale_color_manual(values = c(
    "Young" = "blue",
    "Middle" = "purple",
    "Old" = "red"
  )) +
  coord_cartesian(
    ylim = c(0, max(nsyb_data$value, na.rm = TRUE) * 1.35)
  ) +
  labs(
    x = NULL,
    y = "MitoTimer Red:Green Fluorescence Ratio"
  ) +
  theme_classic() +
  theme(
    legend.position = "none"
  )

ggsave("NSYB.png",p_nsyb,
       width = 6, height = 4, dpi = 300)
p_nsyb
```


```{r create plots for Repo red:green across age groups}
p_repo <- ggplot(repo_data, aes(x = age, y = value, color = age)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, color = "gray50") +
  geom_jitter(width = 0.05, height = 0, size = 2.5, alpha = 0.65) +
  stat_pvalue_manual(
    repo_stats,
    label = "p.adj.signif",
    tip.length = 0.01,
    hide.ns = FALSE
  ) +
  scale_color_manual(values = c(
    "Young" = "blue",
    "Middle" = "purple",
    "Old" = "red2"
  )) +
  coord_cartesian(ylim = c(0, 0.65)) +
  labs(
    x = NULL,
    y = "MitoTimer Red:Green Fluorescence Ratio"
  ) +
  theme_classic() +
  theme(legend.position = "none")

ggsave("REPO.png",p_repo_zoom,
       width = 6, height = 4, dpi = 300)

p_repo

```


```{r merge plots}
p_repo_zoom <- p_repo +
  coord_cartesian(ylim = c(0, 0.65))

ggsave("mitotimer_age_panels.png", p_nsyb + p_repo_zoom,
       width = 10, height = 4, dpi = 300)

p_nsyb + p_repo_zoom
```

## Data Analysis for Control 10-15 day Paraquat Treated Flies

### Clean Data for Analysis 

```{r cleaning names for control data}
control_10_15 <- data_long %>%
  filter(grepl("10to15", summary)) %>%
  mutate(
    genotype = case_when(
      grepl("nSyb", summary) ~ "nSyb",
      grepl("Repo", summary) ~ "repo"
    ),
    treatment_group = case_when(
      grepl("NegativeControl", summary) ~ "No Treatment",
      grepl("PositiveControl", summary) ~ "Paraquat"
    ),
    genotype = factor(genotype, levels = c("nSyb", "repo")),
    treatment_group = factor(treatment_group, levels = c("No Treatment", "Paraquat"))
  )
```

```{r check tables}
control_stats <- control_10_15 %>%
  group_by(genotype) %>%
  t_test(value ~ treatment_group) %>%
  add_xy_position(x = "genotype", dodge = 0.8)
control_stats
control_stats <- control_stats %>%
  mutate(
    p_label = case_when(
      p <= 0.0001 ~ "****",
      p <= 0.001  ~ "***",
      p <= 0.01   ~ "**",
      p <= 0.05   ~ "*",
      TRUE        ~ "ns"
    )
  )
```

### Run Stats on Control Data
```{r run stats on control data}
control_stats

save_stats <- function(df, filename) {

  df_clean <- df %>%
    select(where(~ !is.list(.)))

  write.csv(
    df_clean,
    filename,
    row.names = FALSE
  )
}

save_stats(
  control_stats,
  "results/control_10_15_ttest.csv"
)

```
### Plot Control Data

```{r run stats and plot control 10-15 day data}
p_control_10_15 <- ggplot(
  control_10_15,
  aes(x = genotype, y = value)
) +
  
  geom_boxplot(
    aes(group = interaction(genotype, treatment_group)),
    position = position_dodge(width = 0.8),
    width = 0.45,
    outlier.shape = NA,
    fill = "white",
    color = "gray55"
  ) +
  
  geom_point(
    aes(fill = treatment_group),
    shape = 21,
    color = "black",
    size = 2.7,
    alpha = 0.85,
    stroke = 0.8,
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.8
    )
  ) +
  
  stat_pvalue_manual(
    control_stats,
    label = "p_label",
    tip.length = 0.01,
    hide.ns = FALSE
  ) +
  
  scale_fill_manual(values = c(
    "No Treatment" = "white",
    "Paraquat" = "black",
    "control" = "white",
    "paraquat" = "black"
  )) +
  
  coord_cartesian(
    ylim = c(0, max(control_10_15$value, na.rm = TRUE) * 1.3)
  ) +
  
  labs(
    x = "Genotype",
    y = "MitoTimer Red:Green Fluorescence Ratio",
    fill = "Treatment"
  ) +
  
  theme_classic() +
  theme(
    legend.position = "right"
  )

p_control_10_15

ggsave(
  "control_10_15_plot.png",
  plot = p_control_10_15,
  width = 6,
  height = 4,
  dpi = 300
)

```

## Data Analysis for Intergrated density nSyb and Repo Young, Middle, Old Flies 

### Create function to plot and run stats on intergrated densisty data 

```{r neuron_intergrated_density_data}

make_connected_int_plot <- function(data, group_pattern, plot_title = NULL) {
  
  plot_data <- data %>%
    filter(grepl(group_pattern, summary)) %>%
    select(
      fly_id,
      summary,
      age,
      left_green_int_den,
      right_green_int_den,
      left_red_int_den,
      right_red_int_den
    ) %>%
    distinct() %>%
    pivot_longer(
      cols = c(
        left_green_int_den,
        right_green_int_den,
        left_red_int_den,
        right_red_int_den
      ),
      names_to = "roi_channel",
      values_to = "int_density"
    ) %>%
    mutate(
      int_density = as.numeric(gsub(",", "", int_density)),
      
      age = case_when(
        grepl("Young", age, ignore.case = TRUE) ~ "Young",
        grepl("Middle", age, ignore.case = TRUE) ~ "Middle",
        grepl("Old", age, ignore.case = TRUE) ~ "Old",
        TRUE ~ as.character(age)
      ),
      age = factor(age, levels = c("Young", "Middle", "Old")),
      age_num = as.numeric(age),
      
      channel = case_when(
        grepl("green", roi_channel, ignore.case = TRUE) ~ "Green",
        grepl("red", roi_channel, ignore.case = TRUE) ~ "Red"
      ),
      channel = factor(channel, levels = c("Green", "Red")),
      
      side = case_when(
        grepl("left", roi_channel, ignore.case = TRUE) ~ "Left",
        grepl("right", roi_channel, ignore.case = TRUE) ~ "Right"
      ),
      
      sample_id = paste(fly_id, summary, side, sep = "_"),
      
      x_pos = case_when(
        channel == "Green" ~ age_num - 0.18,
        channel == "Red" ~ age_num + 0.18
      )
    ) %>%
    group_by(sample_id, age, age_num, channel, x_pos) %>%
    summarise(
      int_density = mean(int_density, na.rm = TRUE),
      .groups = "drop"
    )
  
  line_data <- plot_data %>%
    select(sample_id, age, channel, x_pos, int_density) %>%
    pivot_wider(
      names_from = channel,
      values_from = c(x_pos, int_density)
    ) %>%
    drop_na(
      x_pos_Green,
      x_pos_Red,
      int_density_Green,
      int_density_Red
    )
  
  ggplot(plot_data, aes(x = x_pos, y = int_density)) +
    
    geom_segment(
      data = line_data,
      aes(
        x = x_pos_Green,
        xend = x_pos_Red,
        y = int_density_Green,
        yend = int_density_Red
      ),
      inherit.aes = FALSE,
      color = "gray75",
      alpha = 0.65,
      linewidth = 0.40
    ) +
    
    geom_violin(
      aes(group = interaction(age, channel), fill = channel),
      trim = FALSE,
      alpha = 0.28,
      width = 0.32,
      color = "gray30",
      linewidth = 0.6,
      bounds = c(0, Inf)
    ) +
    
    geom_point(
      aes(color = channel),
      size = 2.0,
      alpha = 0.80
    ) +
    
    stat_summary(
      aes(group = interaction(age, channel)),
      fun = mean,
      geom = "point",
      color = "black",
      fill = "black",
      shape = 21,
      size = 2.5,
      stroke = 0.6
    ) +
    
    scale_fill_manual(values = c(
      "Green" = "forestgreen",
      "Red" = "firebrick2"
    )) +
    
    scale_color_manual(values = c(
      "Green" = "limegreen",
      "Red" = "red2"
    )) +
    
    scale_x_continuous(
      breaks = c(1, 2, 3),
      labels = c("Young", "Middle", "Old")
    ) +
    
    scale_y_continuous(
      breaks = c(
        0,
        25000000,
        50000000,
        75000000,
        100000000,
        125000000
      ),
      labels = c(
        "0",
        "2.5e7",
        "5.0e7",
        "7.5e7",
        "1.0e8",
        "1.25e8"
      ),
      expand = expansion(mult = c(0.02, 0.05))
    ) +
    
    labs(
      title = plot_title,
      x = NULL,
      y = "Integrated Density",
      fill = "Channel",
      color = "Channel"
    ) +
    
    theme_classic(base_size = 16) +
    
    theme(
      plot.title = element_text(
        size = 20,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 14)
      ),
      
      axis.title.x = element_blank(),
      axis.title.y = element_text(
        size = 18,
        color = "black",
        face = "bold",
        margin = margin(r = 10)
      ),
      
      axis.text.x = element_text(
        size = 15,
        color = "black",
        face = "bold"
      ),
      
      axis.text.y = element_text(
        size = 13,
        color = "black"
      ),
      
      legend.title = element_text(
        size = 15,
        face = "bold"
      ),
      
      legend.text = element_text(
        size = 14
      ),
      
      legend.position = "right",
      
      plot.margin = margin(15, 25, 15, 15)
    )
}


```

## Run stats on Integrated density Data 
```{r save inden data}

# Save the data used inside each integrated density plot
nsyb_int_data <- p_nsyb_red_green_int$data
repo_int_data <- p_repo_red_green_int$data

write.csv(
  nsyb_int_data,
  "results/nsyb_integrated_density_data.csv",
  row.names = FALSE
)

write.csv(
  repo_int_data,
  "results/repo_integrated_density_data.csv",
  row.names = FALSE
)

# Run and save stats: Red vs Green within each age group
nsyb_int_stats <- nsyb_int_data %>%
  group_by(age) %>%
  t_test(int_density ~ channel, paired = TRUE)

repo_int_stats <- repo_int_data %>%
  group_by(age) %>%
  t_test(int_density ~ channel, paired = TRUE)

nsyb_int_stats_save <- nsyb_int_stats %>%
  select(where(~ !is.list(.)))

repo_int_stats_save <- repo_int_stats %>%
  select(where(~ !is.list(.)))

write.csv(
  nsyb_int_stats_save,
  "results/nsyb_integrated_density_red_vs_green_ttest.csv",
  row.names = FALSE
)

write.csv(
  repo_int_stats_save,
  "results/repo_integrated_density_red_vs_green_ttest.csv",
  row.names = FALSE
)

nsyb_int_stats
repo_int_stats

```

### Chart Integrated Density Plots

```{r}
p_nsyb_red_green_int <- make_connected_int_plot(
  data_long,
  "nSyb_Young|nSyb_Middle|nSyb_Old",
  NULL
)

p_repo_red_green_int <- make_connected_int_plot(
  data_long,
  "Repo_Young|Repo_Middle|Repo_Old",
  NULL
)

p_nsyb_red_green_int
p_repo_red_green_int


```

### Save Integrated Density Plots

```{r}


ggsave(
  "nsyb_absolute_integrated_density.png",
  plot = p_nsyb_red_green_int,
  width = 7,
  height = 5.5,
  dpi = 600
)

ggsave(
  "repo_absolute_integrated_density.png",
  plot = p_repo_red_green_int,
  width = 7,
  height = 5.5,
  dpi = 600
)


```
